---
title: "PSYC 301 Exam Formulas"
subtitle: "(scroll down for the exam you need)"
author: "Patrick Bolger, PhD"
date: "`r format(Sys.time(), '%d %B %Y')`"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
output: 
  rmdformats::readthedown:
    use_bookdown: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F, comment = NA )
```

```{r SuppressScientificNotation}
options(scipen=999)
```


```{r}

pacman::p_load(readr, DT, dplyr)

```


# Exam 1

$$\overline{x}=\frac{1}{N}\sum_{i=1}^N x_i$$

<br/>

$$s=\sqrt{\frac{\sum(x_i-\overline{x})^2}{N-1}}$$
<br/>

$$s^2=\frac{\sum(x_i-\overline{x})^2}{N-1}$$

<br/>

$$z_i=\frac{(x_i-\overline{x})}{s}$$

# New to Exam 2

$$x_i=z_i(s)+\overline{x}$$

<br/>

$$SE_{mean}=\frac{\sigma}{\sqrt{N}}\ or\ \frac{SD}{\sqrt{N}}$$

<br/>

$$CI_{95\%}=\overline{X}\pm \bigg(1.96 \times \frac{\sigma}{\sqrt{N}}\bigg)$$

<br/>

$$CI_{99\%}=\overline{X}\pm \bigg(2.58 \times \frac{\sigma}{\sqrt{N}}\bigg)$$

# New to Exam 3

<br/>

## $\chi^2$

$$\chi_{goodness\ of\ fit}^2=\sum_{i=1}^k \frac{(O_i-E_i)^2}{E_i}$$

<br/>

## Standard Error of the mean (repeated from above)

$$SE_{mean}=\frac{\sigma}{\sqrt{n}}$$

<br/>

## *z*-test (not the *z*-score, but rather the test)

$$z=\frac{\bar{x}-\mu_0}{SE_{mean}}$$
<br/>

or

$$z=\frac{\bar{x}-\mu_0}{\sigma/\sqrt{n}}$$
<br/>

## One-sample *t*-test

$$t=\frac{\bar{x}-\mu_0}{\hat{\sigma}/\sqrt{n}}$$
or

$$t=\frac{\bar{x}-\mu_0}{s/\sqrt{n}}$$

## Student's *t*-test

$$t=\frac{\overline{x}_1-\overline{x}_2}{\sqrt{\left[\frac{(n_1-1)s_1^2+(n_2-1)s_2^2}{n_1+n_2-2}\right]\left[\frac{n_1+n_2}{n_1n_2}\right]}}$$
<br/>

## Cohen's *d*

$$d=\frac{\bar{x}_1-\bar{x}_2}{\sqrt{\frac{s_1^2+s_2^2}{2}}}$$

## *F*-ratio for a Oneway ANOVA

$$F=\frac{Mean\ Squares_{between}}{Mean\ Squares_{within}}$$

<br/>


$$=\frac{Sum\ of\ Squares_{between}}{degrees\ of\ freedom_{between}}
\bigg/\frac{Sum\ of\ Squares_{within}}{degrees\ of\ freedom_{within}}$$

<br/>

$$=\frac{\sum\limits_{k=1}^G N_k(\overline{Y}_k-\overline{Y})^2}{G-1}\Bigg/\frac{\sum\limits_{k=1}^G\sum\limits_{i=1}^{N_k}(Y_{ik}-\overline{Y}_k)^2}{N-G}$$

<br/>

### $\eta^2$

$$\eta^2=\frac{Sum\ of\ Squares_{between}}{Sum\ of\ Squares_{total}}$$

<br/>

## ANOVA table

| **source of variance** | **degrees of freedom** | **sum of squares** | **mean squares**   | ***F***-**statistic** | ***p***-**value** |
|--------------------|--------------------|----------------|----------------|---------------|-----------|
| **between groups**     | $df_b$     | $SS_b$ | $MS_b$ | $F$             | $p$         |
| **within groups**      | $df_w$      | $SS_w$  | $MS_w$  | -             | -         |
| **total**              | $df_t$       | $SS_t$   | -              | -             | -         |

<br/>

or

| **source of variance** | **degrees of freedom** | **sum of squares** | **mean squares**   | ***F***-**statistic** | ***p***-**value** |
|--------------------|--------------------|----------------|----------------|---------------|-----------|
| **between groups**     | $G-1$     | $\sum\limits_{k=1}^G N_k(\overline{Y}_k-\overline{Y})^2$ | $MS_b=\frac{SS_b}{df_b}$ | $F=\frac{MS_b}{MS_w}$             | *complicated formula*         |
| **within groups**      | $N-G$      | $\sum\limits_{k=1}^G\sum\limits_{i=1}^{N_k}(Y_{ik}-\overline{Y}_k)^2$  | $MS_w=\frac{SS_w}{df_w}$  | -             | -         |
| **total**              | $df_b+df_w$       | $SS_b+SS_w$   | -              | -             | -         |

<br/>

## Pearson's *r*

<br/>

### using *z*-scores

$$r_{x,y}=\frac{\sum{z_{x_i}z_{y_i}}}{n-1}$$

<br/>

### using raw scores only

$$r_{x,y}=\frac{n\Sigma XY-\Sigma X \Sigma Y}{\sqrt{[n\Sigma X^2-(\Sigma X)^2][n\Sigma Y^2-(\Sigma Y)^2]}}$$

<br/>

## Simple regression

<br/>

### slope

$$b=\frac{\sum{XY}-\Big(\big[\sum{X}\sum{Y}\big]/n\Big)}{\sum{X^2}-\Big[(\sum{X)^2}/n\Big]}$$

<br/>

### intercept

$$a=\frac{\sum{Y}-b\sum{X}}{n}$$

<br/>

### residual

$$y_i-\hat{y}$$

# New to Final Exam

<br/>

## Paired-samples *t*-test

$$t=\frac{\sum{D}}{\sqrt{\frac{n\sum{D^2}-(\sum{D})^2}{n-1}}}$$
<br/>

which turns out to be the same as the following;
<br/>

$$t=\frac{\bar{D}}{s_D\big/\sqrt{n}}$$

### Cohen's $d_z$

$$Cohen's\ d_z=\frac{\bar{x}_D}{s_D}$$

<br/>

which is the same as the following:

$$Cohen's\ d_z=\frac{\bar{D}}{\hat{\sigma}_D}$$

---

<br/><br/><br/>

# Normal (*z*-) tables

```{r importZ, include=F}

# raw file shows area to the left of specific z-scores
ztable_leftward <- round(read_csv("tables/ztable.csv", col_types = cols()),4)
# remove lines that are all zeros
ztable_leftward_filtered <- ztable_leftward %>% 
  filter(if_all(starts_with("0"), ~ . > 0)) %>%
  filter(if_all(starts_with("0"), ~ . < 1)) 
ztable_rightward_filtered <- ztable_leftward_filtered %>% mutate(across(`0`:`0.09`, ~ 1-.x))

ztable_leftward_filtered <- ztable_leftward_filtered %>% mutate(across(`0`:`0.09`, ~ round(.x,4)))
ztable_rightward_filtered <- ztable_rightward_filtered %>% mutate(across(`0`:`0.09`, ~ round(.x,4)))

```

```{r zContainer}

zsketch <- htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'First 2 digits of z-score'),
      th(colspan = 10, '3rd digit of z-score'),
    ),
    tr(lapply(
      c('0.00', '0.01', '0.02', '0.03', '0.04', 
        '0.05', '0.06', '0.07', '0.08', '0.09'), th)
      )
    )
  )
)

```

## *z*-table (rightward)

**Directions**: To get any particular *z*-score, combine the value in the row in the left-most column (`First 2 digits of z-score`) with the appropriate value across the top of the rightward columns (`3rd digit of z-score`). For example, the area `to the right` of a *z*-score of `-1.25` would be found by finding the intersection of `-1.2` in the first column with the column labeled `0.05` across the top (This value is 0.8944).

<br/>

```{r zDisplayRightward}

DT::datatable(
  ztable_rightward_filtered, 
  container = zsketch,
  rownames = FALSE,
  extensions = 'KeyTable',
  options = list(keys = TRUE)
  )

```
<br/>

## *z*-table (leftward)

**Directions**: To get any particular *z*-score, combine the value in the row in the left-most column (`First 2 digits of z-score`) with the appropriate value across the top of the rightward columns (`3rd digit of z-score`). For example, the area `to the left` of a *z*-score of `-1.25` would be found by finding the intersection of `-1.2` in the first column with the column labeled `0.05` across the top (this value is 0.1056).

<br/>

```{r zDisplayLeftward}

DT::datatable(
  ztable_leftward_filtered, 
  container = zsketch,
  rownames = FALSE,
  extensions = 'KeyTable',
  options = list(keys = TRUE),
  caption = 
  )

```

---

<br/>

# Critical Values

## $\chi^2$ table

```{r importX2}

x2table <- round(read_csv("tables/x2table.csv", col_types = cols()),4)

```

```{r x2Display}

x2sketch <- htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'degrees of freedom'),
      th(colspan = 3, 'Alpha Level'),
    ),
    tr(lapply(
      c('.05', '.01', '.001'), th)
      )
    )
  )
)

DT::datatable(
  x2table[,c(1,3,5,6)], 
  container = x2sketch,
  rownames = FALSE,
  extensions = 'KeyTable',
  options = list(keys = TRUE),
  caption = 
  )

```

<br/>

## *t*-table

```{r importT}

ttable <- round(read_csv("tables/ttable.csv", col_types = cols()),4)

```

### two-tailed

```{r displayT2tail}

tsketch2tail <- htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'degrees of freedom'),
      th(colspan = 2, 'Alpha level'),
    ),
    tr(lapply(
      c('.05', '.01'), th)
      )
    )
  )
)

DT::datatable(
  ttable[,c(1,5,7)], 
  container = tsketch2tail,
  rownames = FALSE,
  extensions = 'KeyTable',
  options = list(keys = TRUE),
  caption = htmltools::tags$caption(
    style = 'caption-side: bottom; text-align: left;',
    'The absolute bottommost row in this entire table represents infinite
    degrees of freedom, or the Normal curve.')
  )

```

<br/>

### one-tailed

```{r displayT1tail}

tsketch1tail <- htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'degrees of freedom'),
      th(colspan = 2, 'Alpha level'),
    ),
    tr(lapply(
      c('.05', '.01'), th)
      )
    )
  )
)

DT::datatable(
  ttable[,c(1,4,6)], 
  container = tsketch1tail,
  rownames = FALSE,
  extensions = 'KeyTable',
  options = list(keys = TRUE),
    caption = htmltools::tags$caption(
    style = 'caption-side: bottom; text-align: left;',
    'The absolute bottommost row in this entire table represents infinite degrees of freedom, or the Normal curve.')
  )

```

<br/>

## *F*-table ($\alpha=.05$)

```{r importF05}

ftable05 <- round(read_csv("tables/ftable05.csv", col_types = cols()),4)

```



```{r f05Display}

Fsketch <- htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'df within'),
      th(colspan = 18, 'degrees of freedom between'),
    ),
    tr(lapply(
      c('1', '2', '3', '4', '5', '6', '7', '8', '9',
        '10', '12', '15', '20', '24', '30', '40', '60', '120'), th)
      )
    )
  )
)

DT::datatable(
  ftable05, 
  container = Fsketch,
  rownames = FALSE,
  extensions = 'KeyTable',
  options = list(keys = TRUE),
  caption = 
  )

```

<br/>

## *F*-table ($\alpha=.01$)

```{r importF01}

ftable01 <- round(read_csv("tables/ftable01.csv", col_types = cols()),4)

```



```{r f01Display}

DT::datatable(
  ftable01, 
  container = Fsketch,
  rownames = FALSE,
  extensions = 'KeyTable',
  options = list(keys = TRUE),
  caption = 
  )

```

<br/>


## table for Pearson's *r*

### one-tailed

```{r importR}

rtable <- round(read_csv("tables/rtable.csv", col_types = cols()),4)

```



```{r rDisplay1tailed}

Rsketch1tailed <- htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'degrees of freedom'),
      th(colspan = 3, 'Alpha Level'),
    ),
    tr(lapply(
      c('.05', '.025', '.005'), th)
      )
    )
  )
)

DT::datatable(
  rtable, 
  container = Rsketch1tailed,
  rownames = FALSE,
  extensions = 'KeyTable',
  options = list(keys = TRUE),
  caption = 
  )

```

<br/>

### two-tailed

```{r rDisplay2tailed}

Rsketch2tailed <- htmltools::withTags(table(
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'degrees of freedom'),
      th(colspan = 3, 'Alpha Level'),
    ),
    tr(lapply(
      c('.10', '.05', '.01'), th)
      )
    )
  )
)

DT::datatable(
  rtable, 
  container = Rsketch2tailed,
  rownames = FALSE,
  extensions = 'KeyTable',
  options = list(keys = TRUE),
  caption = 
  )

```

<br/>